---
title: "Economic impact of presidents and war"
author: "Spencer Graves and Jouni Helske"
date: "`r Sys.Date()`"
output:
      rmarkdown::html_vignette:
        fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Average Income Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}{inputenc}
---
```{r setup, include = FALSE, cache = FALSE}
library(RefManageR)
bibFile <- system.file("Bib", "biblatexExamples.bib", 
                           package = "RefManageR")
bib <- ReadBib(bibFile, check = FALSE)
bibF2 <- "AvgInc.bib"
bib2 <- ReadBib(bibF2) 
BibOptions(check.entries = FALSE, style = "markdown", cite.style = "authoryear",
           bib.style = "numeric")
```

# Abstract 

This vignette illustrates the use of the [KFAS](https://rweb.crmda.ku.edu/cran/web/packages/KFAS/KFAS.pdf) package for R to evaluate the claim that [World War II (WW II)](https://en.wikipedia.org/wiki/Great_Depression#World_War_II_and_recovery), not [Franklin Roosevelt (FDR)](https://en.wikipedia.org/wiki/Franklin_D._Roosevelt), ended the [Great Depression](https://en.wikipedia.org/wiki/Great_Depression).  This is tne dominant view of the evolution of the US economy between 1933 and 1945.  It is, however, superficially contradicted by our naive review of a plot of average annual income in the US ([real US Gross Domestic Product (GDP) per capita](https://en.wikipedia.org/wiki/Real_gross_domestic_product), 1790-2014, from [MeasuringWorth](http://measuringworth.com/)) in two ways:  First, other wars have not been accompanied by such obvious growth spurts.  Second, FDR's economic performance even without World War II seems unmatched by any other U.S. president.  This document describes alternative attempts to quantify the economic impact of wars separate from that of the administration ("president").  This includes a destailed discussion and comparison of alternative models of economic growth and how [KFAS](https://rweb.crmda.ku.edu/cran/web/packages/KFAS/KFAS.pdf) was used to fit them.  We have two conclusions:  First, KFAS provides a powerful set of tools for modeling extensions to the present analysis, e.g., mixing data from multiple sources such as joint models of GDP per capita with unemployment and mixing annual and quarterly data.  Second, any conclusions regarding FDR vs. WW II must be considered tentative, pending further research using other data.  

# Introduction  

The Legacy of the [New Deal](https://en.wikipedia.org/wiki/New_Deal) and the [Great Depression](https://en.wikipedia.org/wiki/Great_Depression) still carry profound implactions for the current economic policies of the world's most economically advanced countries.  `r Citet(bib2, "Horwitz2011")` recently claimed that Hoover had more agressive deficit spending and public works programs than Franklin Roosavelt, and far from contributing to the recovery, these programs actually made the economy worse.  He concluded that "The persistence of the Hoover myth continues to justify the counterproductive policies of the Obama administration and thereby prevents markets from generating the economic recovery of which they are fully capable."  

Whatever the specific cause, the growth of the U.S. economy under FDR before the war was not matched by that of any other president (not counting the negative record of [Herert Hoover](https://en.wikipedia.org/wiki/Herbert_Hoover)).  This document describes alternative attempts to quantify the economic impact of wars separate from that of the administration ("president").  This includes a detailed discussion and comparison of alternative models of economic growth and how [KFAS](https://rweb.crmda.ku.edu/cran/web/packages/KFAS/KFAS.pdf) was used to fit them.  The conclusions regarding FDR vs. WW II must be considered tentative, pending further research using other data.  

# Plots

We start by plotting the [MeasuringWorth](http://measuringworth.com/) estimates of average annual income in the US (adjusted for inflation) from 1790 to 2014:  

```{r,fig.cap = "real US GDP per capita (thousands of 2009 $)"}
library(Ecdat)
plot(realGDPperCapita/1000~Year, USGDPpresidents)     
```     

This plot could be changed in many ways to make it more informative.  First, this [data.frame](https://stat.ethz.ch/R-manual/R-devel/library/base/html/data.frame.html) covers many years for which we have no GDP data.  Let's start by checking missing data:  

```{r}
which(GDPna <- is.na(USGDPpresidents$realGDPperCapita))
```

The only NAs are the first 33 observations.  Those contain data on other variables for years prior to 1790.  We won't use any of that data for this analysis.  Let's discard it:  

```{r}
GDP <- USGDPpresidents[!GDPna,]
```

Let's try that plot again, this time on a log scale with improved labeling:  

```{r,fig.cap = "real US GDP per capita (thousands of 2009 $)"}
plot(realGDPperCapita/1000~Year, GDP, log='y', type='l',
     ylab='thousands of 2009 dollars', las=1, 
     main='Real US GDP per capita')     
```     

This plot suggests relatively steady growth from 1790 to 2014 with the exception of a dramatic rise roughly two thirds of the way through the series preceeded and followed by declines.  

The president is identified in an [ordered factor](https://stat.ethz.ch/R-manual/R-devel/library/base/html/factor.html), "executive", giving the names of the chief executive (president since 1790).  We need to drop the unused levels from the colonial period:  

```{r}
GDP$executive <- ordered(GDP$executive)
levels(GDP$executive)
```

This [data.frame](https://stat.ethz.ch/R-manual/R-devel/library/base/html/data.frame.html) includes three variables for war.  The first is an [ordered factor](https://stat.ethz.ch/R-manual/R-devel/library/base/html/factor.html).  We need to drop the unused levels of "war", as we just did for "executive":  

```{r}
GDP$war <- ordered(GDP$war)
levels(GDP$war)
```

The next war variable is the number of "battleDeaths" for each war apportioned by the number of days in each year between the official start and end of each war.  However, it hardly makes sense to use this directly as a measure of the impact of the war on the economy, as the nation grew from 3.9 million souls in 1790 to 319 million in 2014.  Instead, we use battle deaths per million population:  

```{r,fig.cap = "Battle deaths per million population"}
plot(battleDeathsPMP~Year, GDP, type='l',
     ylab='battle deaths per million pop', las=1, 
     main='Battle deaths during war')     
```     

Wars involving fewer than 10 battle deaths per million US population per year are not included.  No war since Vietnam has reache that threshold.  This explains why the 2001-2014 [War in Afghanistan](https://en.wikipedia.org/wiki/War_in_Afghanistan_(2001%E2%80%9314)) and the 2003-2011 [Iraq war](https://en.wikipedia.org/wiki/Iraq_War) do not appear here:  As noted in the help page for USGDPpresidents, the two together averaged fewer than 3 battle deaths per year per million population.  

This document first adds annotations to the above plot of realGDPperCapita then considers a series of models that estimate a base growth rate separate from other potential explanitory variables such as the administration (the president) and wars.  Readers interested in the results are encouraged to focus on the plots and tables, skipping the math and software.  

# Graphical analysis 

To search for the exceptional period in the above plot we will compute the change from each year to the next in log(real GDP per capita).  However, before doing that, we first reduce this [data.frame](https://stat.ethz.ch/R-manual/R-devel/library/base/html/cumsum.html) to the period for which realGDPperCapita is available:  

```{r}
sel <- !is.na(USGDPpresidents$realGDPperCapita)
range(USGDPpresidents$Year[sel])
USGDPpres <- USGDPpresidents[sel,]
```

Because "executive" is an [ordered factor](https://stat.ethz.ch/R-manual/R-devel/library/base/html/cumsum.html), we will manually drop the unused levels:  

```{r}
USGDPpres$executive <- ordered(USGDPpres$executive)
```

To evaluate the war effect, the [data.frame](https://stat.ethz.ch/R-manual/R-devel/library/base/html/data.frame.html) USGDPpresidents in the Ecdat package inlcudes a "war" factor identifying all the major wars in US history;  for this purpose, "major war" is defined as averaging more than 10 battle deaths per million population per year and having appeared in the [Correlates of War](https://en.wikipedia.org/wiki/Correlates_of_War) data for conflicts since 1816 or the Wikipedia ["List of wars involving the United States"](https://en.wikipedia.org/wiki/List_of_wars_involving_the_United_States) for wars between 1610 and 1816.  

We want to drop the unused levels for "war" as for "executive":  

```{r}
USGDPpres$war <- ordered(USGDPpres$war)
```

This process identified nine major wars in this subset of USGDPpresidents:  

```{r}
(wars <- levels(USGDPpres$war))
nWars <- length(wars)
```

We also convert UDGDPpres$realGDPperCapita to a time series, because some of the software we will use functions better with the time information in that format:

```{r}
USGDPpres$realGDPperCapita <- with(USGDPpres, 
          ts(realGDPperCapita, Year[1]))
```

Now let's compute the years where the GDP differed most from the previous year:  

```{r}
difGDP <- diff(log(USGDPpres$realGDPperCapita))
```

A run up or down *ends* where difGDP changes sign from one year to the next:  

```{r}
chgSgn <- ((head(difGDP, -1)*tail(difGDP, -1))<0)
```

We convert this logical vector to an index numbering the runs using [cumsum](https://stat.ethz.ch/R-manual/R-devel/library/base/html/cumsum.html):  

```{r}
GDPrun <- c(1, 1+cumsum(chgSgn))
```

We use [tapply](https://stat.ethz.ch/R-manual/R-devel/library/base/html/tapply.html) to identify the beginning and end of each run and to compute run length and total and average growth:  

```{r}
yr0 <- head(USGDPpres$Year, -1)
runStart <- tapply(yr0, GDPrun, min)
yr1 <- tail(USGDPpres$Year, -1)
runEnd <- tapply(yr1, GDPrun, max)
nRuns <- length(difGDP)
runLength <- tapply(rep(1, nRuns), GDPrun, sum)
totGrowth <- tapply(difGDP, GDPrun, sum)
avgGrowth <- tapply(difGDP, GDPrun, mean)
```

We organize these numbers into a [data.frame](https://stat.ethz.ch/R-manual/R-devel/library/base/html/data.frame.html) for convenience:  

```{r}
str(GDPruns <- data.frame(runStart=runStart, 
      runEnd=runEnd, runLength=runLength, 
      totalGrowth=totGrowth, meanGrowth=avgGrowth))
```

To focus on the potentially most interesting periods, we select the runs with averge growth exceeding 0.08 (positive or negative):  

```{r}
GDPruns[abs(GDPruns$meanGrowth)>0.08,]
```

This gives us six runs, three of which occur in the period that jumps off the above plot at us:  1929-1933, 1933-1937, and 1938-1944.  The first period is the presidency of [Herbert Hoover](https://en.wikipedia.org/wiki/Herbert_Hoover), and the last two encompass the [FDR](https://en.wikipedia.org/wiki/Franklin_D._Roosevelt) years of the [Great Depression](https://en.wikipedia.org/wiki/Great_Depression) and [World War II](https://en.wikipedia.org/wiki/World_War_II).  (The break in the middle is the "[Recession of 1937-38](https://en.wikipedia.org/wiki/Recession_of_1937%E2%80%9338)", which convinced FDR to restore the high levels of government spending that seemed to make the difference between the disaster of 1929-33 and the growth of 1934-36.)

We label these points on the above plot:  
     
```{r,fig.cap = "Presidents & exceptional change in GDP"}
with(USGDPpres, 
  plot(realGDPperCapita/1000, log='y', 
       main=paste0('US average annual income\n', 
                         '(GDP per capita)'), 
       ylab='thousands of 2009 dollars', las=1, 
       xlab='Year') )
abline(v=c(1929, 1933, 1945), lty='dashed', col='grey')
text(1930, 2.5, "Hoover", srt=90, cex=0.9)
text(1939.5, 30, 'FDR', srt=90, cex=1.1, col='blue')
```

(This plot requires a slightly different syntax than the previous plot, because the standard [plot](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/plot.html) function won't accept the standard syntax when plotting an object of class "ts" as of 2016-02-09.)

Let's create a function to label these wars with Hoover and FDR on a plot:  

```{r}
plotPresWars <- function(x, ..., 
     yHoover=0.8, yFDR=2, yWar=seq(1.7, by=-.1, len=9), 
     cex.war=0.67, xlab='Year'){
  plot(x, xlab=xlab, ...)
  abline(v=c(1929, 1933, 1945), lty='dashed')
  usr <- par('usr')
  Hy <- (usr[3]+yHoover*diff(usr[3:4]))
  if(par('ylog'))Hy <- exp(Hy)
  text(1930, Hy, 'Hoover', srt=90, cex=.9, xpd=NA)
  Ry <- (usr[3]+yFDR*diff(usr[3:4]))
  if(par('ylog'))Ry <- exp(Ry)
  text(1939, Ry, 'FDR', srt=90, cex=1.1, col='blue', xpd=NA)
  yrMid <- mean(usr[1:2])
  dyr <- diff(usr[1:2])
  y. <- seq(usr[3], usr[4], len=5)
  dy <- diff(y.[c(2,4)])
  y2.5 <- mean(y.[c(2,3)])
  for(i in 2:nWars){
    w <- wars[i]
    sel <- (USGDPpres$war==w)
    yrs <- range(USGDPpres$Year[sel])
    abline(v=yrs, lty='dotted', col='grey')
    yr. <- mean(yrs)
    wy <- (y2.5 + yWar[i-1]*dy)
    if(par('ylog'))wy <- exp(wy)
    text(yr., wy, w, srt=90, col='red', cex=cex.war, 
         xpd=NA)
  }
  invisible("done")
}
```

Now try it:  

```{r,fig.cap = "Presidents, wars & change in GDP"}
with(USGDPpres, 
  plotPresWars(realGDPperCapita/1000, log='y', 
             main=paste0('US average annual income\n', 
                         '(GDP per capita)'), 
             ylab='thousands of 2009 dollars', las=1) )
```

The rest of this vignette describes building various models of economic growth including especially the impact of changes of administration and wars.  First we outline our basic approach to modeling these kinds of data.  

# Bayesian sequential updating:  overview  

The models used here are all state space models, which we discuss in general here, giving specific examples later.  State space modeling is also known as Kalman (Kalman filtering, Kalman smoothing), and dynamic (linear) modeling.  For present purposes, we use the [KFAS](https://rweb.crmda.ku.edu/cran/web/packages/KFAS/KFAS.pdf) package, though many other R packages provide some support for this class of models.  

The theory behind these models can be described in terms of a 3-step Bayesian upadating cycle to add new observations to a probability distribution summarizing the available knowledge about the state of the system under study `r AutoCite(bib2, "Graves07")`:

  1.  Receive and store new data.  
  
  1.  Create a prior distribution for the new data by modifying the posterior computed after the last observations to allow for possible changes in state.  
  
  1.  Combine the new data with the prior to produce a posterior distribution of the state.  
  
In the present discussion, the system under study is the US ecomomy at various times, which we will summmarize in a vector of one or more numbers that represent the state of the economy at a particular point in time.  The term "state" in "state space" refers to this type of representation.)  The data will be annual observations on either real US GDP per capita (realGDPperCapita) or inflation (GPDdeflator) or a vector of both.  

The first model of this type considered here will considers univariate observations $y_t$ = the real GDP per capita on a hidden univariate state $\alpha_t$, which represents the portion of the annual number that persists.  In other words, the differences between the observations and predictions based on the hidden state are assumed to be uncorrelated across time;  any serial correlation in $y_t$ is assumed to be due to the hidden state,  $\alpha_t$.  

The general upward trend in the above plots of real GDP per capita indicate that a model with a univariate state will not fit very well. We use it to fix ideas and to serve as a basis for evaluating the importance of estimating the growth rate at time $t$ separate from the level -- and for evaluating the general utility of this approach to modeling time series.  

With univariate observations on a univariate state, the result is almost a traditional [exponentially weighted moving average (EWMA)](https://en.wikipedia.org/wiki/Exponential_smoothing), except that the weight on the last observation is adjusted with each observation consistent with the Bayesian sequential updating cycle outlined above, converging to an asymptote determined by the ratio of the migration and noise variances `r AutoCite(bib2, "Graves02")`.  

In the more general [Kalman filtering](https://en.wikipedia.org/wiki/Kalman_filter) terminology, the weight on the last observation (or the wieght on the deviation of that observation from prediction) is called the Kalman gain.  Bayesian sequential updating produces a varying Kalman gain, though in many applications the Kalman gain converges to a constant.  Before fitting any model, we first describe the general [KFAS](https://rweb.crmda.ku.edu/cran/web/packages/KFAS/KFAS.pdf) model and show how the Bayesian EWMA described above is a special case.  This formulation allows the system to vary over time even to the extent of supporting asynchronous observations on different variables at irregular points in time.  

The next generalization beyond an EWMA is [double exponential smoothing](https://en.wikipedia.org/wiki/Exponential_smoothing#Double_exponential_smoothing):  It has a two-dimensional state vector with the first dimension being the level and the second being the rate of growth.  The plots above suggests that double exponential smoothing might work relatively well for realGDPperCapita except for 1929-1947.  To model that period, we expand the state vector further by adding random effects for presidents and wars.  

In the following, we shall consider nine different though related models:  

 || Model | state vector components | parameters to estimate
-|------|-------------------------|----------------------
1 | EWMA1: Standard Bayesian EWMA | 1:level | 2:lnQ, lnH
2 | EWMA2: double EWMA | 2:level, slope | 3:lnQ.lvl, lnQ.slope, lnH
3 | EWMA2pres1: EWMA2 allowing slope to change only between presidents | 2:level, slope | 3:lnQ.lvl, lnQ.slope, lnH 
4 | EWMA2pres2: EWMA2 assuming no continuity in slope between presidents | 2:level, slope | 3:lnQ.lvl, lnQ.slope, lnH 
5 | EWMA2pres3: EWMA2 allowing an increase in the slope migration variance between presidents | 2:level, slope | 4:lnQ.lvl, lnQ.slope, lnQ.slopePres, lnH 
6 | EWMA2pres4: EWMA2 with an adjustment for each president | 3:level, slope, Pres | 4:lnQ.lvl, lnQ.slope, lnQ.slopePres, lnH 
7 | EWMA2pres5: EWMA2 with a separate component for each president | 43:level, slope, Pres1, ..., Pres41 | 4:lnQ.lvl, lnQ.slope, lnQ.slopePres, lnH 
8 | EWMA2presWar1: EWMA2pres4 plus war as a fixed effect | 4:level, slope, Pres, War | 4:lnQ.lvl, lnQ.slope, lnQ.slopePres, lnH
9 | EWMA2presWar2: EWMA2pres4 plus war as a random effect | 4:level, slope, Pres, War | 5:lnQ.lvl, lnQ.slope, lnQ.slopePres, lnQ.slopeWar, lnH 

# 1.  The simplest Kalman Filter:  an Exponentially Weighted Moving Average (EWMA)

Consider a univariate noisy measurement, $y_t$, of a univariate state, $\alpha_t$:

$$y_t = \alpha_t + \epsilon_t$$

where the state, $\alpha_t$, follows a random walk:  

$$\alpha_{t+1} = \alpha_t + \eta_t.$$

In this model, we assume that observation error, $\epsilon_t$, and the random migration, $\eta_t$, are normally distributed with mean 0 and variance, $H$ and $Q$, respectively, and the initial state $\alpha_1$ is normal with mean $a_1$ and variance $P_1$.  We shall also sometimes write $a_1$ and $P_1$ as $a_{1|0}$ and $P_{1|0}$ to make it clear that these are the prior mean and variance of $\alpha_1$ before $y_1$ is available (using the conditional subscript notation, following, e.g., `r Citet(bib2, "Harvey")`).  We will then also use $a_{t|t}$ and $P_{t|t}$ to denote the posterior mean and variance given $D_t = \{y_t, y_{t-1}, ...\}$.  The prior distribution for $\alpha_t$ given $D_{t-1}$ and the posterior for $\alpha_t$ given $D_t$ will be computed by recursion as outlined below. (See `r Citet(bib2, "DurbinKoopman2")` for more detail on this model.)

These equations are called the measurement and state transition equations, respectively.  

We will estimate $H$, $Q$, $a_1$ and $P_1$ to maximize the likelihood.  We will also estimate the series $\alpha_t$ conditional on these estimates $H$, $Q$, $a_1$ and $P_1$.  

With this notation, we can provide more details in discussing the 3-step Bayesian sequential updating process outlined above: 

  1.  A new observation, $y_t$, arrives.  
  
  1.  The prior distribution $f(\alpha_t | D_{t-1})$ is computed from the previous posterior $f(\alpha_{t-1} | D_{t-1})$), where $D_{t-1}$ is all the information available prior to the arrival of $y_t$;  for $t$ = 0, this is already assumed as $\alpha_1 \tilde\ N(a_{1|0}, P_{1|0})$.  Otherwise, we have by recursion that the posterior for $(\alpha_{t-1} | D_{t-1})$ is $N(a_{t-1|t-1}, P_{t-1|t-1})$.  From this using the state transition equation, we get the next prior $(\alpha_t | D_{t-1})$ is $N(a_{t|t-1}, P_{t|t-1})$, where $a_{t|t-1} = a_{t-1|t-1}$ and $P_{t|t-1} = P_{t-1|t-1} + Q.$  (The double subscripts help clarify the math, which can be confusing otherwise.  This will be particularly valuable with generalizations discussed below.)  

  1.  Bayes' theorem is then used to update this posterior to incorporate the information in the new observation.  This converts $f(\alpha_t | D_{t-1})$ into $f(\alpha_t | D_t)$.    
  
In our initial application, the observations $y_t$ represent annual observations on real US GDP per capita (realGDPperCapita).  The distribution of the state vector $\alpha_t$ is governed by certain hyperparameters, suppressed in the notation $f(\alpha_t | D_t)$.  These hyperparameters are estimated by maximum likelihood with tests and confidence intervals or regions obtains using standard likelihood ratio procedures.  

# KFAS package for Kalman Filtering and Smoothing

State space modeling with the [KFAS](https://rweb.crmda.ku.edu/cran/web/packages/KFAS/KFAS.pdf) package assumes the following generalization of the above measurement and state transition equations:  

$$y_t = Z_t \alpha_t + \epsilon_t$$

$$\alpha_{t+1} = T_t \alpha_t + R_t \eta_t.$$

where $\epsilon_t \tilde\ N(0, H_t)$, $\eta_t \tilde\ N(0, Q_t)$ and $\alpha_1 \tilde\ N(a_1, P_1)$, idendependent of each other.  In this, $y_t$ is a $p$-vector (where $p$ may actually vary with $t$), $\alpha_t$ is $m$-dimensional, and $\eta_t$ has dimension $k$.  (See `r Citet(bib2, "DurbinKoopman")` for more detail on this model.)   

[KFAS](https://rweb.crmda.ku.edu/cran/web/packages/KFAS/KFAS.pdf) modeling begins with a formula that is converted to an "SSModel" object using the [SSModel](http://rpackages.ianhowson.com/cran/KFAS/man/SSModel.html) function.  This is passed to [fitSSM](http://rpackages.ianhowson.com/cran/KFAS/man/fitSSM.html) to estimate unknown parameters to maximize [logLik.SSModel](http://rpackages.ianhowson.com/cran/KFAS/man/logLik.SSModel.html).  Confidence and prediction intervals can be obtained using [predict.SSModel](http://rpackages.ianhowson.com/cran/KFAS/man/predict.SSModel.html).  Function [KFS](http://rpackages.ianhowson.com/cran/KFAS/man/KFS.html) will produce filtered and smoothed estimates of the state vector in components "a" and "alphahat", respectively, of an object of class "KFS":

<center> 
[formula](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/formula.html) -> [SSModel](http://rpackages.ianhowson.com/cran/KFAS/man/SSModel.html) -> [fitSSM](http://rpackages.ianhowson.com/cran/KFAS/man/fitSSM.html) -> [predict.SSModel](http://rpackages.ianhowson.com/cran/KFAS/man/predict.SSModel.html) -> [KFS](http://rpackages.ianhowson.com/cran/KFAS/man/KFS.html)  
</center>

Before starting to use this machinery, we first note that the [data.frame](https://stat.ethz.ch/R-manual/R-devel/library/base/html/data.frame.html) USGDPpresidents actually includes data back to 1610:

```{r}
range(USGDPpresidents$Year)
```

However, data on realGDPperCapita only begins in 1790:  

```{r}
sel <- !is.na(USGDPpresidents$realGDPperCapita)
range(USGDPpresidents$Year[sel])
```

We now create a formula:  

```{r}
library(KFAS)
EWMA1fmla <- log(realGDPperCapita)~
  SSMtrend(1,Q=list(matrix(NA)), a1=log(realGDPperCapita[1]) )
```

This is converted to an SSModel:  

```{r}
EWMA1mdl <-SSModel(EWMA1fmla, USGDPpres, H=matrix(NA) )
```

This model has 2 knowns, the NA's in EWMA1fmla and SSModel:   

  * Q = the state migration variance  
  
  * H = observation error variance.  

The default behavior of [fitSSM](http://rpackages.ianhowson.com/cran/KFAS/man/fitSSM.html) is to estimate the logarithms of these two variance.  [fitSSM](http://rpackages.ianhowson.com/cran/KFAS/man/fitSSM.html) requires us to provide starting values for log(Q) and log(H), in that order:  

```{r}
EWMA1fit<-fitSSM(EWMA1mdl, inits=c(lnQ=0, lnH=0))
```

One measure of the quality of the fit is the log(likelihood):  

```{r}
logLik(EWMA1fit$model)
```

We compute the estimated migration and observation standard deviations as follows:  

```{r}
exp(EWMA1fit$optim.out$par/2)
```

These estimates put the observation error at essentially zero.  It would be interesting to construct a confidence region for these two parameters, but we shall not attempt that here.  

To help us understand this model, we compute the one-step ahead predictions as follows:  

```{r}
EWMA1pred <- predict(EWMA1fit$model, filtered = TRUE)
```

The one-step ahead prediction errors are as follows:  

```{r}
EWMA1.v <- (log(USGDPpres$realGDPperCapita)-EWMA1pred)         
```

These numbers are actually computed by [KFS](http://rpackages.ianhowson.com/cran/KFAS/man/KFS.html):  '

```{r}
EWMA1.KFS <- KFS(EWMA1fit$model)
range(EWMA1.v-EWMA1.KFS$v)
```

Let's plot this vs. Presidents and wars:  

```{r,fig.cap = "EWMA residuals, Presidents and wars"}
plotPresWars(EWMA1.KFS$v, 
    yHoover=0.8, yFDR=.3, yWar=seq(.6, by=-.1, len=9), 
    ylab='one-step ahead prediction errors')
```

This plot deserves study:  

  1.  During the Hoover administration, the economy performed far worse than this simple Bayesian EWMA predicted.  This is what we expected from earlier macroeconomic research, but it's nice to see it confirmed.  (It increases our confidence in the [KFAS](https://rweb.crmda.ku.edu/cran/web/packages/KFAS/KFAS.pdf) software.)
  
  1.  During the FDR years including WW II, the economy outperformed the EWMA predictions, except for the [Recession of 1937-38](https://en.wikipedia.org/wiki/Recession_of_1937%E2%80%9338).  (Again, this increases our confidence in the software.)  
  
  1.  We do NOT generally see that other wars are accompanied with large positive residuals, as we would expect if WW II and not FDR ended the Great Depression.  There were interesting spikes during the [Northwest Indian War](https://en.wikipedia.org/wiki/Northwest_Indian_War), the [U.S. Civil War](https://en.wikipedia.org/wiki/American_Civil_War), and the [Spanish](https://en.wikipedia.org/wiki/Spanish%E2%80%93American_War)-American-[Philippine](https://en.wikipedia.org/wiki/Philippine%E2%80%93American_War) war.  However, they were isolated and not obviously greater than other spikes that did not accompany wars.  
  
For future reference, we'll compute the root mean square prediction errors for this model:  
  
```{r}
sqrt(mean(EWMA1.v^2))
```  
  
Because log(1+x) is approximately x when x is small, we can interpret this as saying that 0.046 indicates a root mean square prediction error of 4.6 percent.    
  
Let's now consider a double EWMA, which is a similar model that estimates both level and slope.  

# 2.  Double exponential smoothing:  More realistic but still not great for the Great Depression 

[Double exponential smoothing](https://en.wikipedia.org/wiki/Exponential_smoothing#Double_exponential_smoothing) runs an EWMA on the slope of a line and uses that in adjusting the level.  The standard Bayesian modification of that adjusts the weight on the last observation in proportion to the information content of the last observation relative to information accumulated in the previous estimate of the state, as suggested above.  This can be written using the [KFAS](https://rweb.crmda.ku.edu/cran/web/packages/KFAS/KFAS.pdf) notation above as follows:  

$$y_t = Z_t \alpha_t + \epsilon_t$$

$$\alpha_{t+1} = T_t \alpha_t + R_t \eta_t.$$

where $y_t$ is the observation at time $t$ on log(readGDPperCapita), as before, 

  $\alpha_t$ has two components corresponding to the level and slope,
  
  $Z_t$ is a row vector of (1, 0), so $y_t$ is the estimated level plus noise, 
  
  $T_t$ is a 2 x 2 matrix with 0 in the lower left and 1's elsewhere, so the new level is the previous level plus the slope while the new slope is the previous slope, plus migration noise:  
  
$$T_t = \left[\begin{array}
{rrr}
1 & 1 \\
0 & 1 
\end{array}\right]
$$  

Similar to the simple EWMA considered above, $H_t$ is the observation variance.  However, the state vector $\alpha_t$ is 2-dimensional.  To manage this, we assume that $R_t$ is the 2-dimensional identity matrix, and $Q_t$ is a 2-dimensional diagonal matrix.  This gives us 3 parameters to estimate:  The two diagonal elements of the migration variance matrix $Q_t$ and the observation error variance $H_t$.  As before, by default, the algorithm estimated the logarithms of these parameters, which we will call lnQ.level, lnQ.slope, and lnH.  

Modeling is a straightforward generalization of the above following the same process:

<center> 
[formula](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/formula.html) -> [SSModel](http://rpackages.ianhowson.com/cran/KFAS/man/SSModel.html) -> [fitSSM](http://rpackages.ianhowson.com/cran/KFAS/man/fitSSM.html) -> [KFS](http://rpackages.ianhowson.com/cran/KFAS/man/KFS.html)  
</center>

We start by supplying a formula using [SSMtrend(2, ...)](http://rpackages.ianhowson.com/cran/KFAS/man/SSModel.html).  We select a starting value for a1 with 0 slope, because that will make the EWMA1 discussed above a special case of this model (with zero migration variance for the slope).  That's important, because it allows us to test the statistical significance of this additional parameter using [Wilks's theorem](https://en.wikipedia.org/wiki/Likelihood-ratio_test#Distribution:_Wilks.27s_theorem), which says that twice the log(likelihood ratio) of nested models is approximately distributed as chi-square with degrees of freedom equal to number of parameters in the larger model fixed to create the smaller model.  `r Citet(bib2, "PinBates")` note that Wilk's theorem requires the true but unknown parameter values to be in the interior of the parameter space, and testing for a zero variance violates that assumption.  They recommend simulation to estimate p-values, but provide an example that suggests that 0.5 times the p-value for a chi-square(1) provides a reasonable approximation for the actual p-value when testing if a single variance parameter is zero;  we use this approximation below.  (They provide other examples that show that this simple rule does not generalize and adjusts in the wrong directly with eliminating fixed effects from a mixed model;  see their pp. 87-89.)

As before, we start by specifying a formula for the double EWMA:  

```{r}
(a1.2 <- c(level=log(USGDPpres$realGDPperCapita[1]), slope=0))
EWMA2fmla <- log(realGDPperCapita)~
  SSMtrend(2,Q=list(lnQ.lvl=matrix(NA), lnQ.slope=matrix(NA)), 
           a1=a1.2)
```

As above, we convert this to an [SSModel](http://rpackages.ianhowson.com/cran/KFAS/man/SSModel.html):  
```{r}
EWMA2mdl <-SSModel(EWMA2fmla, USGDPpres, H=matrix(NA))
```

We pass this as with the simple EWMA to [fitSSM](http://rpackages.ianhowson.com/cran/KFAS/man/fitSSM.html):  

```{r}
EWMA2init <- c(lnQ.level=0, lnQ.slope=0, lnH=0)
EWMA2fit<-fitSSM(EWMA2mdl, inits=EWMA2init)
(EWMA2logLik <- logLik(EWMA2fit$model))
```

We pause here compute 2*log(likelihood ratio) to test the statistical significance of the slope: 

```{r}
EWMA1logLik <- logLik(EWMA1fit$model)
(EWMAlogLik2.1 <- 2*(EWMA2logLik-EWMA1logLik))
```

However, as discussed above, the assumptions for Wilk's theorem do not apply here.  The standard recommendation would be to compute p-value for this using Monte Carlo.  Before going to that trouble, we first try the modification suggested by `r Citet(bib2, "PinBates")`, mentioned above:  

```{r}
0.5*pchisq(EWMAlogLik2.1, 1, lower=FALSE)
```

This gives us 6e-7.  Even if this is off by two or three orders of magnitude, the effect is still statistically significant -- as we would expect it to be just looking at the plots above.  

Let's continue as above by examining the estimated migration and observation error standard deviations:  

```{r}
exp(EWMA2fit$optim.out$par/2)
```

The migration standard deviation for the level is only slightly smaller that for the EWMA1 above.  The migration standard deviation for the slope and measurement noise seem much smaller than we might have expected.  It would be interesting to compute a 95 percent confidence interval on the standard deviation of the slope migration;  we will not attempt there here at the moment;  the preferred approach, as just indicated, would be Monte Carlo.  

We will skip the call to [predict.SSModel](http://rpackages.ianhowson.com/cran/KFAS/man/predict.SSModel.html), because what we want is returned by [KFS](http://rpackages.ianhowson.com/cran/KFAS/man/KFS.html):  

```{r}
EWMA2.KFS <- KFS(EWMA2fit$model)
```

To help us understand this model, we plot the smoothed estimates of the state vector over time:  

```{r,fig.cap = "Estimated level and slope of GDP"}
plotPresWars(EWMA2.KFS$alphahat,   
  yHoover=0.35, yFDR=.7, yWar=seq(.5, by=-.05, len=9), 
  main="EWMA2 smoothed state")
```

The slope has the shape we expected from the plot, but it does not seem to fit the Hoover and FDR years well.  

To confirm this, let's plot the one-step ahead prediction errors, labeling presidents and wars:  

```{r,fig.cap = "Double EWMA, Presidents and wars"}
plotPresWars(EWMA2.KFS$v, 
  yHoover=0.8, yFDR=.3, yWar=seq(.6, by=-.1, len=9), 
  ylab='one-step ahead prediction error')
```

The root mean square prediction error is as follows:  

```{r}
sqrt(mean(EWMA2.KFS$v^2))
``` 

At 0.043, this is only slightly smaller than the root mean square prediction error of 0.046 for the standard Bayesian EWMA discussed above.  

Let's see if we can do better adding a random effect for presidents.  We consider five different ways to do this, as outlined in the table above.   

# 3.  Allowing the slope to change only between executives 

A simple presidential effect model is to assume that the slope is constant within each presidency but can change between administrations.  This can still be done using [KFAS](https://rweb.crmda.ku.edu/cran/web/packages/KFAS/KFAS.pdf) but requires using a state transition variance, Q, that is not constant.  This requires us to also create an "updatefn" for [fitSSM](http://rpackages.ianhowson.com/cran/KFAS/man/fitSSM.html), because the default "updatefn" will not handle a time varying Q:  

<center> 
[(make time varying the migration variance, Q)](http://rpackages.ianhowson.com/cran/KFAS/man/KFAS.html) -> [formula](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/formula.html) -> [SSModel](http://rpackages.ianhowson.com/cran/KFAS/man/SSModel.html) 

-> [(create updatefn)](http://rpackages.ianhowson.com/cran/KFAS/man/fitSSM.html) -> [fitSSM](http://rpackages.ianhowson.com/cran/KFAS/man/fitSSM.html) -> [KFS](http://rpackages.ianhowson.com/cran/KFAS/man/KFS.html)  
</center>

The state transition variance, Q, is typically specified as a list in [SSMtrend](http://rpackages.ianhowson.com/cran/KFAS/man/SSModel.html).  Before changing this, let's first look at the structure of EWMA2mdl:  

```{r}
str(EWMA2mdl)
```

The most important thing to note for present purposes is that the component Q is "num [1:2, 1:2, 1] NA 0 0 NA":

```{r}
EWMA2mdl$Q
```

This was specified above as "Q=list(lnQ.lvl=matrix(NA), lnQ.slope=matrix(NA))" in the call above to [SSMtrend](http://rpackages.ianhowson.com/cran/KFAS/man/SSModel.html) that created EWMA2fmla.  We will change that here by replacing "matrix(NA)" with an array of all zeros except for making the slope component NA for the last year of each administration:  

```{r}
EWMA2pres1Q.slope <- array(0, c(1,1,225))
contin <- duplicated(USGDPpres$executive)
presLastYr <- c(!contin[-1], FALSE)
EWMA2pres1Q.slope[,,presLastYr] <- NA 
```

We now create a formula for this model:  

```{r}
EWMA2pres1fmla <- log(realGDPperCapita) ~ 
  SSMtrend(2, Q = list(matrix(NA), EWMA2pres1Q.slope), a1=a1.2) 
```

As before, we use this to build an [SSModel](http://rpackages.ianhowson.com/cran/KFAS/man/SSModel.html): 
```{r}
EWMA2pres1mdl <-SSModel(EWMA2pres1fmla, USGDPpres, H=matrix(NA))
```

The next step is to create the desired updatefn assuming we estimate three log(variance) parameters for lnQ.level, lnQ.slope, and lnH, as for the double EWMA above:    

```{r}
EWMA2pres1updt <- function(pars=EWMA2init, model=EWMA2pres1mdl, 
                           ...){
  vars <- exp(pars)
  Q <- model$Q
  Q[1,1,] <- vars[1]
  Q[2,2,presLastYr] <- vars[2]
  model$Q <- Q
  model$H[] <- vars[3]
  model
} 
```

We're now ready to call [fitSSM](http://rpackages.ianhowson.com/cran/KFAS/man/fitSSM.html):  

```{r}
EWMA2pres1fit <- fitSSM(EWMA2pres1mdl, inits=EWMA2init, 
                        updatefn=EWMA2pres1updt)
logLik(EWMA2pres1fit$model)
```

This is identical to the log(likelihood) of the double EWMA above.  

As before, we'll compute the smoothed estimate of the state vector and other related quantities:  

```{r}
EWMA2pres1.KFS <- KFS(EWMA2pres1fit$model)
```

Let's check the root mean square error in one-step ahead predictions:  
```{r}
sqrt(mean(EWMA2pres1.KFS$v^2))
``` 

This is 0.0427, matching the previous model, as we expected.  

Let's plot the state vector:  

```{r,fig.cap = "Estimated level and slope of GDP for first President model"}
plotPresWars(EWMA2pres1.KFS$alphahat, 
             yHoover=0.72, yFDR=.3, yWar=seq(.6, by=-.1, len=9), 
             main='Simplest president model')
```

This is very similar to the double EWMA but with the slope held constant within administration -- precisely what we attempted to fit.  
However, since the model seems not to improve upon the double EWMA, we will leave this here and try the second model mentioned above.  

# 4.  Assuming no continuity in slope between executives 

Another approach to modeling a president effect is to restart the slope from 0 at the beginning of each presidency.  We will do this by modifying the EWMA2pres1 model above so slope component of the state transition matrix is 0 for the last year of each presidency and 1 otherwise:  

$$T_t = \left[\begin{array}
{rrr}
1 & 1 \\
0 & \gamma_t  
\end{array}\right], 
$$  

where $\gamma_t$ = 0 for the last year of a presidency, and 1 otherwise.  

We'll accomplish this by adding another step to the process used above, modifying the [SSModel](http://rpackages.ianhowson.com/cran/KFAS/man/SSModel.html) before fitting it:  

<center> 
[(make time varying the migration variance, Q)](http://rpackages.ianhowson.com/cran/KFAS/man/KFAS.html) -> [formula](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/formula.html) -> [SSModel](http://rpackages.ianhowson.com/cran/KFAS/man/SSModel.html) -> 

[(make time varying the state transition matrix, T)](http://rpackages.ianhowson.com/cran/KFAS/man/KFAS.html) -> [(create updatefn)](http://rpackages.ianhowson.com/cran/KFAS/man/fitSSM.html) -> [fitSSM](http://rpackages.ianhowson.com/cran/KFAS/man/fitSSM.html) -> [KFS](http://rpackages.ianhowson.com/cran/KFAS/man/KFS.html)  
</center>

In this particular case, we can use the same Q and [formula](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/formula.html) as in the previous section, and modify the previous model to make T time varying as desired:   

```{r}
EWMA2pres2mdl <- EWMA2pres1mdl
T.other <- EWMA2pres1mdl$T
N <- NROW(EWMA2pres1mdl$y)
EWMA2pres2.T <- array(T.other, c(2,2,N), 
      dimnames=dimnames(T.other) )
EWMA2pres2.T[2,2,presLastYr] <- 0
EWMA2pres2mdl$T <- EWMA2pres2.T
```

We're not quite done modifying this model:  As noted with [help('SSModel')](http://rpackages.ianhowson.com/cran/KFAS/man/SSModel.html), we must set attribute 'tv' to reflect that both T and Q are time varying.  From "str(EWMA2mdl)" above, we see that attr(EWMA2mdl1, "tv")= int [1:5] 0 0 0 0 0, indicating a model in which none of Z, H, T, R, and Q are time varying.  This is different for EWMA2pres1mdl:  

```{r}
attr(EWMA2pres1mdl, 'tv')
```

This says that only Q is time varying.  We just made T time varying but have to change this attribute to make it official.  In the process, we will add names to make the structure easier to read:  

```{r}
attr(EWMA2pres2mdl, 'tv') <- c(Z=0L, H=0L, T=1L, R=0L, Q=1L)
```

Formally at least, we have the same parameters to estimate as with EWMA2pres1;  this allows us to use the same updatefn.  We therefore fit this model as follows:  

```{r}
EWMA2pres2fit <- fitSSM(EWMA2pres2mdl, inits=EWMA2init, 
                        updatefn=EWMA2pres1updt) 
logLik(EWMA2pres2fit$model)
```

This has a slightly better fit, as indicated by the increase in the log(likelihood).  

We next compute the smoothed estimate of the state vector and other related quantities, as before:  

```{r}
EWMA2pres2.KFS <- KFS(EWMA2pres2fit$model)
```

Let's check the root mean square one-step ahead prediction error:  

```{r}
sqrt(mean(EWMA2pres2.KFS$v^2))
```

This also indicates a slightly better fit. 

What does the state vector look like?  

```{r,fig.cap = "Estimated level and slope of GDP, decoupling the presidents"}
plotPresWars(EWMA2pres2.KFS$alphahat, 
             yHoover=0.72, yFDR=.3, yWar=seq(.6, by=-.08, len=9),
             main='Decoupled presidents')
```

The slope here shows substantially more variability.  Let's look at it more closely:  

```{r,fig.cap = "Estimated level slope of GDP, decoupling the presidents"}
plotPresWars(EWMA2pres2.KFS$alphahat[, 'slope'], 
             yHoover=0.72, yFDR=.3, yWar=seq(.6, by=-.08, len=9), 
             ylab='slope')
abline(h=0)
```

This strongly suggests that the Hoover administration was a statistical outlier, while FDR (with World War II) may or may not have been.  To investigate this further, let's look at the slope and it's variance for each president.  [help('KFS')](http://rpackages.ianhowson.com/cran/KFAS/man/KFS.html) tells us that the "V" component is the covariance of "alphahat", given all the data (and the "P" and "Pinf" components are the non-diffuse and diffuse parts of the error covariance of the one-step ahead predicted states in "a").  Just to make sure we understand this correctly, let's plot "V":  

```{r,fig.cap = "Variance of the estimated slope, decoupling the presidents"}
plotPresWars(USGDPpres$Year, EWMA2pres2.KFS$V[2,2,], 
             yHoover=0.72, yFDR=.3, yWar=seq(.6, by=-.08, len=9), 
             type='l', ylab='smoothed slope variance')
```

This is plausibly constant within administration, which we would expect from specifying zero migration variance for the slope.  If this is correct, we would expect the "information" (i.e., resiprocal variance) to be proportional to the number of years in each presidency.  To confirm that, we compute one posterior variance and the number of years for each presidency:    

```{r,fig.cap = "Slope, variance, and number of years for each presidency"}
EWMA2pres2V <- tapply(EWMA2pres2.KFS$V[2,2,], 
                      USGDPpres$executive, mean)
EWMA2pres2n <- table(USGDPpres$executive)
plot(as.numeric(EWMA2pres2n), 1/EWMA2pres2V, 
     xlab='Number of years', ylab='info = 1/(smoothed variance)')
```

This looks like a straight line except for two points.  Who are those two?  

```{r,fig.cap = "Posterior variance vs. number of years for each presidency, labeled"}
plot(as.numeric(EWMA2pres2n), 1/EWMA2pres2V, 
     xlab='Number of years', ylab='info = 1/(smoothed variance)')
text(as.numeric(EWMA2pres2n), 1/EWMA2pres2V, 1:41)
```

Of course:  The two off the line are the first and last.  

Let's compute the slope and confidence limits:  

```{r,fig.cap = "Decoupled slope for each presidency"}
EWMA2pres2pres <- tapply(EWMA2pres2.KFS$alphahat[, 'slope'], 
                         USGDPpres$executive, mean)
sz2 <- qnorm(.975)*sqrt(EWMA2pres2V)
sz2. <- sz2 %o% c(lower=-1, upper=1)
EWMA2pres2slim <- as.numeric(EWMA2pres2pres) + sz2.
EWMA2pres2Slim <- cbind(data.frame(
executive=ordered(names(sz2), names(sz2)), 
slope=EWMA2pres2pres), EWMA2pres2slim)
library(ggplot2)
ggplot(EWMA2pres2Slim, aes(x=executive, y=slope))+geom_point() +
    coord_cartesian(ylim = c(-0.05, 0.08)) + 
    geom_errorbar(ymin=EWMA2pres2Slim$lower,
        ymax=EWMA2pres2Slim$upper)+ 
    geom_hline(aes(yintercept=mean(slope),color="red")) # mean slope
```  

This clearly says that the Hoover and FDR administrations were different from all others, but it still doesn't say whether FDR himself was different or World War II created the difference.  Before considering that, let's continue with the other alternative presidency models.  

# 5.  Allowing the migration variance to increase between executives 

This uses EWMA2pres1mdl from above but estimates one more parameter, and therefore requires a different 'updatefn':   

```{r}
EWMA2pres3init <- c(lnQ.level=0, lnQ.slope=0, 
      lnQ.slopePres=0, lnH=0)
EWMA2pres3updt <- function(pars=EWMA2pres3init, 
                      model=EWMA2pres1mdl, ...){
  vars <- exp(pars)
  Q <- model$Q
  Q[1,1,] <- vars[1]
  Q[2,2,] <- vars[2]
  Q[2,2,presLastYr] <- vars[3]
  model$Q <- Q
  model$H[] <- vars[4]
  model
} 
EWMA2pres3fit <- fitSSM(EWMA2pres1mdl, inits=EWMA2pres3init, 
                        updatefn=EWMA2pres3updt) 
logLik(EWMA2pres3fit$model)
```

This has the same number for log(likelihood) as two of the previous formulations.  Let's check the smoothed state:  

```{r,fig.cap = "Smoothed state increasing migration varaince between presidents"}
EWMA2pres3.KFS <- KFS(EWMA2pres3fit$model)
plotPresWars(EWMA2pres3.KFS$alphahat, 
             yHoover=0.72, yFDR=.3, yWar=seq(.6, by=-.08, len=9) )
```

Officially, this is different from EWMA2pres1.  However, the migration variance within administration is so small, it's indistinguisable from 0.  Let's compute the root mean square one-step ahead prediction error:  

```{r}
sqrt(mean(EWMA2pres3.KFS$v^2))
```

The same number as before, as expected.  

# 6.  EWMA2 with an adjustment for each president 

This model considers a 3-dimensional state vector:  level, slope and a Presidential adjustment for slope.  For this, we will want a state transition matrix to look like the following:  

$$T_t = \left[\begin{array}
{rrr}
1 & 1 & 1 \\
0 & 1 & 0 \\
0 & 0 & \gamma_t 
\end{array}\right],
$$ 

where $\gamma_t$ = 0 for the last year of a presidency, and 1 otherwise, as with model EWMA2pres2mdl, above.  

We will create this by modifying a triple EWMA following a process similar to the above:  

<center> 
[(make time varying the migration variance, Q)](http://rpackages.ianhowson.com/cran/KFAS/man/KFAS.html) -> [formula](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/formula.html) -> [SSModel](http://rpackages.ianhowson.com/cran/KFAS/man/SSModel.html) -> 

[(make time varying the state transition matrix, T)](http://rpackages.ianhowson.com/cran/KFAS/man/KFAS.html) -> [(create updatefn)](http://rpackages.ianhowson.com/cran/KFAS/man/fitSSM.html) -> [fitSSM](http://rpackages.ianhowson.com/cran/KFAS/man/fitSSM.html) -> [KFS](http://rpackages.ianhowson.com/cran/KFAS/man/KFS.html)  
</center>

Let's start by modifying the formula, similar to what we did with EWMA2pres1fmla, above, but for the new 3-dimensional state    

```{r}
a1.3 <- c(level=log(USGDPpres$realGDPperCapita[1]), slope=0, 
           slopePres=0)
EWMA2pres4fmla <- log(realGDPperCapita) ~ 
  SSMtrend(3, Q = list(level=matrix(NA), slope=matrix(NA),
                       slopePres=EWMA2pres1Q.slope),
           a1=a1.3) 
```

We next create an [SSModel](http://rpackages.ianhowson.com/cran/KFAS/man/SSModel.html), as before:   

```{r}
EWMA2pres4mdl <- SSModel(EWMA2pres4fmla, USGDPpres, H=matrix(NA))
```

Now we need to make T time varying.  

```{r}
pres4states <- c('level', 'slope', 'slopePres')
pres4T <- array(EWMA2pres4mdl$T, dim=c(3,3,N), 
        dimnames=list(pres4states, pres4states) )
pres4T[1, 3,] <- 1
pres4T[2, 3,] <- 0 
pres4T[3,3, EWMA2pres4mdl$T] <- 0
EWMA2pres4mdl$T <- pres4T 
```

As for EWMA2pres2mdl, we need also to fix the 'tv' attribute:  

```{r}
attr(EWMA2pres4mdl, 'tv') <- c(Z=0L, H=0L, T=1L, R=0L, Q=1L)
```

We have 4 parameters to estimate, as in the previous model, but the state vector now has 3 dimensions, not 2 as before.  This required a different 'updatefn':   

```{r}
EWMA2pres4updt <- function(pars=EWMA2pres3init, 
                      model=EWMA2pres4mdl, ...){
  vars <- exp(pars)
  Q <- model$Q
  Q[1,1,] <- vars[1]
  Q[2,2,] <- vars[2]
  Q[3,3,presLastYr] <- vars[3]
  model$Q <- Q
  model$H[] <- vars[4]
  model
} 
EWMA2pres4fit <- fitSSM(EWMA2pres4mdl, inits=EWMA2pres3init, 
                        updatefn=EWMA2pres4updt) 
logLik(EWMA2pres4fit$model)
```

This is worse than any other model considered so far except the simplest Bayesian EWMA, which assumed zero slope.  Let's also check the root mean square one-step ahead prediction error:  

```{r}
EWMA2pres4.KFS <- KFS(EWMA2pres4fit$model)
sqrt(mean(EWMA2pres4.KFS$v^2))
```

This confirms the deficiencies of this model fit.  

Let's try another optimizer:  [SSModel](http://rpackages.ianhowson.com/cran/KFAS/man/SSModel.html) calls [optim]((https://stat.ethz.ch/R-manual/R-devel/library/stats/html/optim.html).  [optimx](https://cran.r-project.org/web/packages/optimx/optimx.pdf) use a more modern and robust algorithm.  

To use this, we modify EWMA2pres4updt to return (-logLik(model)):  

```{r}
EWMA2pres4updtx <- function(pars=EWMA2pres3init, 
                      model=EWMA2pres4mdl, ...){
  vars <- exp(pars)
  Q <- model$Q
  Q[1,1,] <- vars[1]
  Q[2,2,] <- vars[2]
  Q[3,3,presLastYr] <- vars[3]
  model$Q <- Q
  model$H[] <- vars[4]
  (-logLik(model))
} 
library(optimx)
EWMA2pres4updtx()
(EWMA2pres4fitx <- optimx(EWMA2pres3init, EWMA2pres4updtx) )
```

This claimed convergence to essentially the same place as [fitSSM(EWMA2pres4mdl, ...)](http://rpackages.ianhowson.com/cran/KFAS/man/fitSSM.html) above.  

I also tried nloptr but couldn't get it to work.  

#```{r}
library(nloptr)
(EWMA2pres4fito <- nloptr(EWMA2pres3init, EWMA2pres4updtx, 
                          model=EWMA2pres4mdl) )
#```

# 7.  EWMA2 with a separate component of the state vector for each president 

[Let's skip this for now -- and maybe delete it later?  ]

# 8 EWMA2pres3 or EWMA2pres4 plus war as a fixed effect

[Which one?]

# 9. EWMA2pres3 or EWMA2pres4 plus war as a random effect

[Which one?]


# Discussion 

The following table summarizes the log(likelihood) and the root mean square error or deviation (rmse) from the one-step ahead prediction for each of the nine models described above with a brief comment on each fit:  

 || Model | state vector components | parameters to estimate | logLik | rmse | comments 
-|------|-------------------------|-----------------|---|---|------
1 | EWMA1: Standard Bayesian EWMA | 1:level | 2:lnQ, lnH | 372.9323 | 0.046 | obviously need slope 
2 | EWMA2: double EWMA | 2:level, slope | 3:lnQ.lvl, lnQ.slope, lnH | 384.6971 | 0.0427 | slope too stiff 
3 | EWMA2pres1: EWMA2 allowing slope to change only between presidents | 2:level, slope | 3:lnQ.lvl, lnQ.slope, lnH | 384.6971 | 0.0427 | same as EWMA2 
4 | EWMA2pres2: EWMA2 assuming no continuity in slope between presidents | 2:level, slope | 3:lnQ.lvl, lnQ.slope, lnH | 389.576 | 0.0421 | slightly better than previous models 
5 | EWMA2pres3: EWMA2 allowing an increase in the slope migration variance between presidents | 2:level, slope | 4:lnQ.lvl, lnQ.slope, lnQ.slopePres, lnH | 384.6971 | 0.0427
6 | EWMA2pres4: EWMA2 with an adjustment for each president | 3:level, slope, Pres | 4:lnQ.lvl, lnQ.slope, lnQ.slopePres, lnH |  382.4992 | 0.0428
7 | EWMA2pres5: EWMA2 with a separate component for each president | 43:level, slope, Pres1, ..., Pres41 | 4:lnQ.lvl, lnQ.slope, lnQ.slopePres, lnH | (not fit) | (not fit)
8 | EWMA2presWar1: EWMA2pres4 plus war as a fixed effect | 4:level, slope, Pres, War | 4:lnQ.lvl, lnQ.slope, lnQ.slopePres, lnH
9 | EWMA2presWar2: EWMA2pres4 plus war as a random effect | 4:level, slope, Pres, War | 5:lnQ.lvl, lnQ.slope, lnQ.slopePres, lnQ.slopeWar, lnH 





# References
```{r results = "asis", echo = FALSE}
PrintBibliography(bib2)
```

# Bibtest 

THIS SECTION TO BE DELETED AFTER THE MARKUP LANGUAGE STYLE IS LEARNED 

This is an example of a citation in the text `r Citet(bib, "loh")`. Now we cite in parentheses `r AutoCite(bib, "baez/online", before = "e.g., ")`.  You can change the default options in a setup chunk at the start of the document or at any other point using the <code>BibOptions</code> function or by specifying options as a list in the `.opts` argument to the cite functions.  In this example we mix `"authoyear"` citation style with `"numeric"` bibliography style.

Note that I do not only have to cite by key, and may use all the features of the `SearchBib` function to index into the BibEntry object.  Here are all the entries of type `Report` in my bibliography `r Citet(bib, bibtype = "Report", .opts = list(hyperlink = "to.doc"))`.  The hyperlinks will take you to their entry in the bibliography.  The link for `r TextCite(bib, "markey")` will open the document in a new window; this is the default behaviour, if a link is available (see `?open.BibEntry`). The following citation has no hyperlink `r AutoCite(bib, location = "Uppsala", .opts = list(hyperlink = FALSE))`.  

  You can change the default options in a setup chunk at the start of the document or at any other point using the <code>BibOptions</code> function or by specifying options as a list in the `.opts` argument to the cite functions.  In this example we mix `"authoyear"` citation style with `"numeric"` bibliography style.

Note that I do not only have to cite by key, and may use all the features of the `SearchBib` function to index into the BibEntry object.  

hyperlinks will take you to their entry in the bibliography.  The link for `r TextCite(bib, "markey")` will open the document in a new window; this is the default behaviour, if a link is available (see `?open.BibEntry`). The following citation has no hyperlink `r AutoCite(bib, location = "Uppsala", .opts = list(hyperlink = FALSE))`.  

Bartholomew59


# Testbib to be deleted after ref style is learned
```{r results = "asis", echo = FALSE}
PrintBibliography(bib, .opts = list(check.entries = FALSE, sorting = "ynt"))
```




